---
output: 
  html_document: 
    keep_md: yes
---

# Cell lineage and division analysis

### 1. Make a video of the cumulative pattern of cell divisions plotted on the tissue

We want to colour code the movie by cell generation. The `tm sm make_db` command will first build a database from the segmented data (or return "Nothing to be done" if the database is present). The next command on the right will run the cell generation analysis and make the video we want. It outputs the data in the output_analysis folder within the same movie directory.

* Copy-paste the following commands in the terminal:

```
tm sm make_db; tm cell_division_generation_pattern.R . output_analysis
```

```{r cumulative_cell_division_pattern, warning=FALSE, message=FALSE, fig.width=11, cache=FALSE, eval=TRUE, echo=FALSE, results='hide'}
movieDir="/Users/retourna/MovieDB/demo"

scriptsDir=Sys.getenv("TM_HOME")
if(is.na(file.info(scriptsDir)$isdir)){
  stop(paste("TM_HOME not correctly defined (",scriptsDir ,")"))
}

source(file.path(scriptsDir, "commons/TMCommons.R"))
source(file.path(scriptsDir, "commons/BaseQueryFunctions.R"))
source(file.path(scriptsDir, "config/default_config.R"))

db <- openMovieDb(movieDir)

# limit generation to a more reasonable range
genColors =c("0"="white", "1" = "red", "2"="green", "3"="cyan", ">3"="magenta")

# Send a SQL query to get the cell lineage
cellsWithLin <- dbGetQuery(db, "select cell_id, lineage_group, generation 
                                from cell_histories") %>%
  # fix a generation cut off for a more reasonable range
  mutate(generation_cutoff=ifelse(generation>3, ">3", ac(generation))) %>%
  # add cell vertices for cell rendering as polygons
  dt.merge(locload(file.path(movieDir, "cellshapes.RData")), by = "cell_id") %>%
  arrange(frame, cell_id, bond_order) %>% print_head()

cellsWithLin %>%
  render_frame(70) + 
  geom_polygon(aes(x_pos, y_pos, fill=as.factor(generation_cutoff), group=cell_id), alpha=0.5) +
  scale_fill_manual(name="generation", values=genColors) +
  ggtitle("Cell_division_generation")

```

[Where to find the results ?](../tm_qs_example_data.md#4-look-at-the-results) **|** 
[Back to tutorial list](../tm_qs_example_data.md#3-select-the-analysis-you-are-interested-in)

### 2. Make a video of coarse-grained cell division nematics plotted on the tissue

We want to calculate the orientation of cell divisions averaged in each element of a square grid, and we want plot the orientation axes (unit nematics) on the tissue. The `tm sm make_db` command will first build a database from the segmented data (or return "Nothing to be done" if the database is already present). The next command then runs the cell division orientation analysis and makes the video we want. It outputs the data in the output_analysis folder within the same movie directory.


* Copy-paste the following commands in the terminal:

```
tm sm make_db; tm cell_division_orientation_pattern.R . output_analysis
```

```{r cell_division_nematic_pattern, warning=FALSE, message=FALSE, fig.width=11, cache=FALSE, eval=TRUE, echo=FALSE, results='hide'}

mqf_cg_grid_unit_nematics_CD(movieDir, gridSize = 90, kernSize = 5) %>%
  ## Plot CD nematics on image #55
  render_frame(7) + 
  geom_segment(aes(x=x1, y=y1, xend=x2, yend=y2),
               size=2, alpha=0.7, lineend="round", color="orange", na.rm=T) +
  ggtitle("Coarse-grained cell division nemtatics")


```

[Where to find the results ?](../tm_qs_example_data.md#4-look-at-the-results) **|** 
[Back to tutorial list](../tm_qs_example_data.md#3-select-the-analysis-you-are-interested-in)


### 3. Plot cell division rate

Now, we want to make a graph that shows the average division rate per cell. The `tm sm make_db` command builds the database if not yet present (or returns "Nothing to be done" if the database is present). The next command does the plot and saves it in the output_analysis folder within the same movie directory.

* Copy-paste the following commands in the terminal:

```
tm sm make_db; tm cell_division_rate.R . output_analysis
```

```{r cell_division_rate, warning=FALSE, message=FALSE, fig.width=7, cache=FALSE, eval=TRUE, echo=FALSE,results='hide'}
source(file.path(scriptsDir, "commons/TimeFunctions.R"))

CDrateByTimeIntervals <- mqf_cg_roi_rate_CD(movieDir, rois = "raw") %>%
  chunk_time_into_intervals(3600) %>%
  group_by(movie, roi,interval_mid) %>%
  summarise(avgCDrate=mean(cell_loss_rate),
            semCD=se(cell_loss_rate),
            time_sec=interval_mid[1],
            dev_time=mean(dev_time))

ggplot(CDrateByTimeIntervals, aes(dev_time, avgCDrate, color=movie)) + 
  geom_line()+
  geom_point(size=1, color="black") +
  geom_errorbar(aes(ymin=(avgCDrate-semCD), ymax=(avgCDrate+semCD)),
                size=0.3, width=0.4, color="black") +
  ylab(expression(paste("CD rate [", cell^-1, h^-1,"]"))) +
  facet_wrap(~roi) +
  ggtitle("Cell_division_rate")

```

[Where to find the results ?](../tm_qs_example_data.md#4-look-at-the-results) **|** 
[Back to tutorial list](../tm_qs_example_data.md#3-select-the-analysis-you-are-interested-in)

### 4. Plot average cell division orientation as a function of time

We now want to plot the average division orientation as a function of time in a circular diagram. The time is color coded.

* Copy-paste the following commands in the terminal:

```
tm sm make_db; tm cell_division_orientation.R . output_analysis
```

```{r cell_division_orientation, warning=FALSE, message=FALSE, fig.width=7, cache=FALSE, eval=TRUE, echo=FALSE,results='hide'}
CDNematics <- mqf_cg_roi_unit_nematics_CD(movieDir, rois = "raw") %>% 
  group_by(movie) %>%
  mutate(maxnormByMovie=max(norm,na.rm=T)) %>% 
  group_by(movie,roi) %>%
  mutate(maxnormByRoi=max(norm,na.rm=T)) %>% print_head() 

ggplot(CDNematics , aes()) + 
  geom_segment(aes(x=phi, y=0, xend=phi, yend=(norm), color=dev_time),size=1, alpha=0.5) +
  geom_segment(aes(x=mod2pi(phi+pi), y=0, xend=mod2pi(phi+pi), yend=(norm),
                   color=dev_time), size=1, alpha=0.5) +
  scale_color_gradientn(name="Time [h]",
                        colours=c("black", "blue", "green", "yellow", "red"),
                        limits=c(min(CDNematics$dev_time),max(CDNematics$dev_time)),
                        na.value = "red") +
  coord_polar(start=-pi/2,direction=+1)+
  scale_x_continuous(breaks=seq(0,3*pi/2,pi/2), 
                     labels=c(expression(pi),expression(paste(pi/2," Ant")),
                              expression(0),expression(-pi/2)),
                     limits=c(0,2*pi)) +
  xlab("") +
  ylab("CD nematic norm") +
  facet_wrap(~roi) +
  ggtitle("avg_CD_nematics")

```

[Where to find the results ?](../tm_qs_example_data.md#4-look-at-the-results) **|** 
[Back to tutorial list](../tm_qs_example_data.md#3-select-the-analysis-you-are-interested-in)

### 5. For further details

* compare multiple movies and ROI's, see [TM R User Manual](https://mpicbg-scicomp.github.io/tissue_miner/user_manual/TM_R-UserManual.html#comparing-averaged-quantities-between-movies-and-rois)
