---
title: 'TissueMiner: an R-tutorial to visualize cell dynamics in 2D tissues'
author: "Raphael Etournay"
date: "25 November 2015"
output:
  html_document:
    fig_caption: yes
    keep_md: yes
    number_sections: yes
    toc: yes
  pdf_document:
    fig_caption: yes
    number_sections: yes
    toc: yes
---



```{r set-options, echo=FALSE, cache=FALSE}
options(width=60)

```


# Introduction to TissueMiner and the R-language

* **Prerequisite:** it is assumed in this tutorial that 
    + time-lapses have been processed using the TissueMiner automated workflow [here](https://github.com/mpicbg-scicomp/tissue_miner)

* **We provide three processed time-lapses** corresponding to the distal part of 3 wild-type *Drosophila* pupal wings. [Please, download them here](https://cloud.mpi-cbg.de/index.php/s/XF5swEw9lxRmZNj). 

* **No programming skill required:** this tutorial will guide the user, without any previous knowledge in programming, through the main aspects of data manipulation and visualization using the R language. 

* **The power of an existing grammar** to manipulate and to visualize data. The open-source R language now provides very efficient and simple ways to handle important amount of tabular data, using a so called "grammar" of data manipulation (*dplyr* package) and visualization (*ggplot2* package) (ref Wickham). We found R to be ideal for querying our relational databases (*RSQLite* package), and to perform subsequent calculations and analyses for understanding tissue morphogenesis. Thus, the user keeps control of his data by learning a few verbs and the simple syntax of the grammar. This gives the advantage of full autonomy on the data that can manipulated without restriction. 

* **TissueMiner extends this grammar** to facilitate the visualization and quantification of cell dynamics during tissue morphogenesis

* **TissueMiner is compatible with the Python programming language**. We provide a Python-tutorial of TissueMiner [here](https://github.com/mpicbg-scicomp/tissue_miner/tree/master/docs)

* **High performance computing platform**. TissueMiner has been designed to be used in command-line in order to easily batch repetitive tasks and to run tasks on a high performance computing platform (cluster). 

* **A comprehensive HOWTO**. This tutorial progressively introduces the techniques of data manipulation and visualization for the user to learn how to master them. It is also organized as an *HOWTO* to visualize and quantify cell dynamics in 2D tissues, by a simple copy-paste of the code encapsulated in shaded grey boxes into a R-shell such as in [RStudio](https://www.rstudio.com/products/rstudio/). We strongly encourage the user to use RStudio or a similar software to conveniently run the R programs provide in TissueMiner.

* **TissueMiner can easily be extended** by the user to address project-specific questions

* **Tutorial layout**: the code is encapsulated in shaded grey boxes that delimit **code chunks**. The **results** are displayed immediately below in an open box in which all lines start with the ## sign. Within code chunks, **comments** are visible in green and are prefixed with at least a # sign. The comments indicate what the following code does. Here is an example:

```{r layout example, warning=FALSE, message=FALSE, echo=T}
# This is a comment: the code below will print "Welcome to TissueMiner"
print("Welcome to TissueMiner")
```

## TissueMiner automated workflow
* Instructions to install and run the automated workflow can be found [here](https://github.com/mpicbg-scicomp/tissue_miner)
* Here is a list of important files that are generated by the automated workflow
    + The database: **(\<movie_name\>.sqlite)**
    + An extra table **(cellshapes.RData)** to represent cell contours by using anticlockwisely ordered cell vertices
    + An extra table **(./roi_bt/lgRoiSmoothed.RData)** to store cells in user-defined regions of interest
    + An extra table **(./topochanges/t1DataFilt.RData)** to store cell neighbor changes
    + An extra table **(./shear_contrib/triangles.RData)** to store triangles
    + Extra tables **(./shear_contrib/\<ROI_name\>/avgDeformTensorsWide.RData)** to store the calculated pure shear deformation of triangles and tissue for each region of interest
        

## Setup TissueMiner routines and Rstudio

### Download TissueMiner routines
#### Method 1
```{bash method 1a, warning=FALSE, message=FALSE, echo=T, eval=F}
# Download TissueMiner (in your home folder):
# go to https://github.com/mpicbg-scicomp/tissue_miner
# click on 'Clone in Desktop' or 'Download ZIP'
```

**Open a terminal and execute the lines below**

```{bash method 1b, warning=FALSE, message=FALSE, echo=T, eval=F}

# Set a path to the tissue_miner folder and export it to the global environement
echo "export TM_HOME=~/tissue_miner" >> .bash_profile
source .bash_profile

```

#### Method 2

**Open a terminal and execute the lines below**

```{bash method 2, warning=FALSE, message=FALSE, echo=T, eval=F}

# Set a path to install TissueMiner in your home folder (please use an absolute path)
echo "export TM_HOME=~/tissue_miner" >> .bash_profile
source .bash_profile

# If Git isn't installed yet, please install it
# On Ubuntu
sudo apt-get install git

# On MacOs
http://git-scm.com/download/mac

# download TissueMiner (require git to be installed)
git clone https://github.com/mpicbg-scicomp/tissue_miner.git ${TM_HOME}

```

### Rstudio installation
Please, follow the instructions [here](https://www.rstudio.com/products/rstudio/) to install RStudio desktop.
```{bash RStudio startup, warning=FALSE, message=FALSE, echo=T, eval=F}

# On MacOSX, setting up the .bash_profile might not be sufficient when starting RStudio from the launcher.
# Just open a terminal (unix-shelll interpreter) and type in:
echo "TM_HOME=~/tissue_miner" >> ~/.Renviron
echo "TM_CONFIG=~/tissue_miner/config/flywing_tm_config.R" >> ~/.Renviron

# On Windows, TM_HOME and TM_CONFIG must be defined in the .Renviron file.
# Jut type in: help("Startup") in a R session (Rstudio), and follow the manual

```

### Avconv installation
```{bash avconv, warning=FALSE, message=FALSE, echo=T, eval=F}
# install avconv (only needed for movie rendering)
# On Ubuntu
sudo apt-get install --assume-yes libav-tools
# On Mac, please visit
http://ffmpegmac.net/
or
http://superuser.com/questions/568464/how-to-install-libav-avconv-on-osx

```

## Load TissueMiner routines in Rstudio
* **Please modify the paths (first chunk below) according to your data location**
* **Always execute the code below before running the analysis**

***

```{r Setup IN/OUT, warning=FALSE, message=FALSE, echo=T,results='hide'}
# Define path to all processed movies: TO BE EDITED BY THE USER
#movieDbBaseDir="/media/junk/Raphael_RawData/ownCloud/MovieRepository_DB"
movieDbBaseDir=Sys.getenv("TUTORIAL_DATA")
# movieDbBaseDir="/Users/retourna/ownCloud/DB"

# Define a working directory where to save the analysis: TO BE EDITED BY THE USER
outDataBaseDir="/tmp/output"
```

***

```{r Setup environment, warning=FALSE, message=FALSE, echo=T,results='hide'}
# Set up path to the TissueMiner code
# This command requires that the global environment TM_HOME is defined in the .bash_profile
scriptsDir=Sys.getenv("TM_HOME")

# CAUTION on MACOS: install xquartz if not yet installed
# http://www.xquartz.org/

# Load TissueMiner libraries
source(file.path(scriptsDir, "commons/TMCommons.R"))
source(file.path(scriptsDir, "db/movie_rotation/RotationFunctions.R"))
source(file.path(scriptsDir, "commons/BaseQueryFunctions.R"))
source(file.path(scriptsDir, "commons/TimeFunctions.R"))
source(file.path(scriptsDir, "config/flywing_tm_config.R"))

# Load a R library
library("zoo")

# Set up working directory
mcdir(outDataBaseDir)

# Set general theme for graphs: more specific tuning must be done for each graph
theme_set(theme_bw())
theme_update(panel.grid.major=element_line(linetype= "dotted", color="black", size=0.2),
             panel.border = element_rect(size=0.3,color="black",fill=NA),
             axis.ticks=element_line(size=0.2),
             axis.ticks.length=unit(0.1,"cm"),
             legend.key = element_blank()
)

# Hardwire isotropic deformation color scheme
isotropColors <- c("division"="orange",
                   "extrusion"="turquoise",
                   "cell_area"="green",
                   "sumContrib"="blue",
                   "tissue_area"="darkred")

# hardwire the movie color scheme
movieColors <- c("WT_1"="blue",
          "WT_2"="darkgreen",
          "WT_3"="red"
)

```

***

## R: the basics

Many books or web sites describe the R language, and we only introduce the necessary knowledge to understand this tutorial.
We recommend of few references that have been useful to us:

* *The art of R programming* by Norman Matloff 
* *R cookbook* by Paul Teetor
* *R graphics cookbook* by Winston Chang
* [https://www.rstudio.com/resources/cheatsheets/](https://www.rstudio.com/resources/cheatsheets/)

### Variable assignment and simple instructions

```{r The basics 1, warning=FALSE, message=FALSE, echo=TRUE, cache=TRUE}
# assign a number to the variables x and y
x <- 2
y <- 3
# display the result of x + y
x + y

# is x equal y?
x==y

# is x different from y?
x!=y

# is x superior to y? 
x>y

# is x inferior to y?
x<y
```

***

### A vector is a series of values. 
```{r The basics 2, warning=FALSE, message=FALSE, echo=TRUE, cache=TRUE}
# assign a vector to x and to y:
x <- c(4,3,2)
y <- c(1,2,3)
# assign a bolean vector to z:
z <- c(TRUE,FALSE,TRUE)
# display the result of x + y (element-wise addition):
x + y
# display the result of x + y + z (z is automatically coerced to integers)
x + y + z
```

***

### Named vectors
In some cases, it is convenient to name each element of the vector. Such a vector is useful to store configuration parameters.
```{r The basics 3, warning=FALSE, message=FALSE, echo=TRUE, cache=TRUE}
# assign a named vector to x:
x <- c("movie1"="red", "movie2"="blue", "movie3"="green")
# display the content of x
x
```

***

### Tabular data: dataframe
Tabular data that we obtain from the relational database are stored in a table refereed to as *dataframe* in the R language. This tutorial essentially shows how to manipulate dataframes in order to perform calculations and prepare the data for plotting. A dataframe is composed of columns that correspond to vectors of identical length.
```{r The basics 4, warning=FALSE, message=FALSE, echo=TRUE, cache=TRUE}
# Assign a data frame to x:
x <- data.frame(frame=c(1,2,3), cell_area=c(20,22,24))
# display the content of x:
x
# display the number of lines in x:
nrow(x)
# display the 2 first rows of x:
head(x, n=2)
# display the 2 last rows of x:
tail(x, n=2)
```

***

## How to query a relational database ?

### Open a connection to the database

* Database format: SQLite
* Open a connection to one database: **openMovieDb()** function using the RSQLite package
* The name of the time-lapse is used to identify the corresponding database
* The SQLite connection is assigned to a variable called "db".


```{r SQLite DB connection, warning=FALSE, message=FALSE, echo=TRUE, cache=FALSE}
# Define path to all time-lapses
# movieDbBaseDir <- "/media/project_raphael@fileserver/movieSegmentation"
# Define path a particular time-lapse called "WT_25deg_111102"
movieDir <- file.path(movieDbBaseDir, c("WT_1"))

# Connection to the DB stored in the "db" variable
db <- openMovieDb(movieDir)

```

```{r SQLite DB disconnection, warning=FALSE, message=FALSE, echo=TRUE, cache=FALSE, eval=FALSE}
# Close the connection
dbDisconnect(db)
```

***

### Query the database using the SQL language

* **Simplicity of the SQL language:** only three words ***select***, ***from***, and ***where*** are sufficient to perform database queries: one can ***select*** the desired columns ***from*** a given table ***where*** the rows of a given column fulfill a user defined criterium.
* A SQL query results in a table or dataframe that we assign to a variable in the R language
* More complicated SQL queries are possible, but we will instead use the grammar of data manipulation provided in R to manipulate the dataframes in the computer memory.

```{r Simple SQL query, warning=FALSE, message=FALSE, echo=TRUE, cache=TRUE, cache.lazy=FALSE}
# Use the built-in "dbGetQuery" function to query the database
# Write SQL statements in quotes
# Assign the resulting data frame to the "cellProperties" variable
cellProperties <- dbGetQuery(db, "select cell_id, frame, area from cells")

# show first lines of the table
head(cellProperties)

# Filter out the margin cell (id 10000) around the tissue
cellProperties <- dbGetQuery(db, "select cell_id, frame, area from cells
                                  where cell_id!=10000") 
```

```{r Simple SQL query select all columns, warning=FALSE, message=FALSE, echo=TRUE, cache=TRUE}
# Select all columns of a table
allCellProperties <- dbGetQuery(db, "select * from cells where cell_id!=10000")
```

```{r MEM CLEAN UP 1, warning=FALSE, message=FALSE, echo=F,results='hide'}
rm(allCellProperties,cellProperties)
```

***

## Manipulate large data sets using a grammar of data manipulation

* Here, we briefly introduce the main verbs and the syntax of the grammar of data manipulation supplied by the dplyr package. In practice, just **a single operator and about 5 verbs only are sufficient to effectively manipulate data**. We also encourage the user to download the Rstudio cheat sheet [here](http://www.rstudio.com/resources/cheatsheets/) in which the grammar is summarized.

* Simply stated, this grammar allows the user to **chain a series of operations** by using the pipe operator **%>%**. In each step of the chain, every intermediate result is taken as an input for the next operation. Each type of operation on dataframes is identified by a verb. 

* This grammar also allows the user to chain other built-in R-functions or custom ones.

In the present tutorial, we mainly use the following few verbs:

Functions | Description | Package
--------------|-------------|-------------
**dbGetQuery** | query a SQLite database and returns a dataframe | RSQLite
**mutate** | perform calculations on columns by adding or modifying existing ones | dplyr
**summarize** | compute summary statistics | dplyr
**group_by** (**ungroup**) | subsets data into chunks prior to a mutate or a summarize operation | dplyr
**filter** | parse data on row content | dplyr
**select** | parse data on column names | dplyr
**arrange** | order values of desired columns | dplyr
**inner_join** | merge two data frames by intersecting user-defined columns | dplyr
**melt** or **gather** | gather columns into rows |reshape2/dplyr
**dcast** or **spread** | spread rows into columns | reshape2/dplyr

***

* This grammar can be easily extended by the user and can be used in combination with important TissueMiner functions for visualizing and quantifying cell dynamics in 2D-living tissues:

Functions | Description | Project
--------------|-------------|-------------
**print_head** | head the current table with the row number | TissueMiner
**dt.merge** | fast merging of two dataframes, possibility to suffix column names | TissueMiner
**openMovieDb** | open a connection to a movie database | TissueMiner
**multi_db_query** | aggregate data into one dataframe | TissueMiner
**coarseGrid** | assign grid elements to cell positions | TissueMiner
**smooth_tissue** | average quantities in time using a moving window | TissueMiner
**align_movie_start** | align movies at earliest common developmental time | TissueMiner
**chunk_time_into_intervals** | undersample time for local time averaging | TissueMiner
**synchronize_frames** | find closest frame to user-defined time intervals | TissueMiner
**mqf_* functions** | set of multi-query functions to quantify cell dynamics | TissueMiner


***

### Learning the grammar on an example
**Aim:** calculate the average cell area in square microns as function of time in hours from start of time-lapse recording. 


**Howto:**

* use the **dbGetQuery()** function to input a dataframe to start the chain of operations
* use the **%>%** operator to chain operations
* manipulate the input dataframe using the *dplyr* grammar 

```{r Example of dplyr, warning=FALSE, message=FALSE, echo=TRUE, cache=TRUE}
# We query the DB to get cell area and pipe the resulting table the next function
avgCellArea <- dbGetQuery(db, "select cell_id, frame, area from cells") %>%
  # remove the huge artificial margin cell around the tissue
  filter(cell_id!=10000) %>%
  # convert pixel to squared microns knowing that 1px = 0.207 micron
  mutate(area_real=(0.207)^2*area) %>%
  # indicate that the next function must be applied frame-wise 
  group_by(frame) %>%
  # calculate the average area in each frame of the time-lapse
  summarize(area_avg=mean(area_real)) %>%
  # cancel grouping
  ungroup() %>%
  # bring time in seconds into the current table by matching the frame number
  inner_join(dbGetQuery(db, "select * from frames"), by="frame") %>%
  # convert time to hours
  mutate(time_h=round(time_sec/3600, 1)) %>%
  # remove the unecessary columns
  select(-c(frame, time_sec)) %>%
  # order time chronologically
  arrange(time_h)
```

***

### Extending the grammar
For convenience, we built a custom **print_head()** function to display the first lines of the current dataframe, without affecting the data content. The **print_head()** function can therefore be placed wherever needed in the chain of operations:

```{r Example of print_head, warning=FALSE, message=FALSE, echo=TRUE, cache=TRUE}
# Here, is again example 1, but we display the first and last steps using print_head()
avgCellArea <- dbGetQuery(db, "select cell_id, frame, area from cells") %>% 
  print_head() %>%
  filter(cell_id!=10000) %>% 
  mutate(area_real=(0.207)^2*area) %>% 
  group_by(frame) %>% 
  summarize(area_avg=mean(area_real)) %>% 
  ungroup() %>% 
  inner_join(dbGetQuery(db, "select * from frames"), by="frame") %>% 
  mutate(time_h=round(time_sec/3600, 1)) %>% 
  select(-c(frame, time_sec)) %>% 
  arrange(time_h) %>% print_head()
```

***

### Vectorized conditional statement (ifelse)
The R language provides a vectorized **ifelse()** function that we can then use in combination with the dplyr grammar. The vectorized **ifelse()** function takes 3 arguments corresponding to the condition (if), the consequent (then), and the alternative (else).

```{r Example of ifelse, warning=FALSE, message=FALSE, echo=TRUE, cache=TRUE}
# Here, is an example in which we display each intermediate step
cell <- dbGetQuery(db, "select cell_id, frame, area, elong_xx, elong_xy from cells") %>% 
  # additional column isMarginCell to flag the margin cell as "true"
  mutate(isMarginCell=ifelse(cell_id==10000, TRUE, FALSE)) %>% print_head()

```

***

### Modify table layout into wide or long formats

#### Wide to long format: the melt() or gather() function.
The  melt() (or gather()) function creates two columns: 

* one 'variable' column listing variable names 
* one 'value' column with their corresponding value.

Both **melt()** and **gather()** are equivalent, gather() being the newest implementation from the dplyr package.

```{r Examples of melt or gather, warning=FALSE, message=FALSE, echo=TRUE, cache=TRUE}
# Example 1: 
# by default, melt() only gathers numerical data into a pair of {variable, value} columns
longFormat <- melt(cell) %>% print_head()
# by default, gather() gathers all columns
longFormat <- gather(cell) %>% print_head()

# Of note, the two columns {cell_id, frame} uniquely define each cell in frame 
# Therefore, to keep consistent data, the frame column should not be gathered

# Example 2: specify which columns to gather into {variable, value} columns
longFormat <- melt(cell, measure.vars = c("area","elong_xx","elong_xy","isMarginCell")) %>%
  print_head()
# Or
longFormat <- gather(cell, variable, value, c(area,elong_xx,elong_xy,isMarginCell)) %>%
  print_head()

# Example 3: specify which columns shouldn't be gathered (equivalent to example 2)
longFormat <- melt(cell, id.vars =  c("cell_id","frame")) %>% print_head()
# Or
longFormat <- gather(cell, variable, value, -c(cell_id,frame)) %>% print_head()


```

***

#### Lond to wide format: the dcast() or spread() function
The  dcast() (or spread()) function creates as many columns as variable names contained in the 'variable' column and lists the corresponding values. Both **dcast()** and **spread()** are equivalent, spread() being the newest implementation from the tidyr package.

```{r Example of dcast or spread, warning=FALSE, message=FALSE, echo=TRUE, cache=TRUE}
# The melt operation is reversible (the row identifiers must be uniquely defined),
# but booleans area coerced into numeric format
# Using dcast(), cell_id and frame are the row identifiers,
# wherease the variable column is spread into column names
example <- cell %>% print_head() %>%
  melt(id.vars =  c("cell_id","frame")) %>% print_head() %>%
  dcast(cell_id+frame~variable, value.var="value") %>% print_head()
# Or
example <- cell %>% print_head() %>%
  gather(variable, value, -c(cell_id,frame)) %>% print_head() %>%
  spread(variable,value) %>% print_head()

```

```{r MEM CLEAN UP 2, warning=FALSE, message=FALSE, echo=F,results='hide'}
rm(cell,longFormat,example)
```

***

## Visualize complex data sets using a grammar of graphics 

* Here, we briefly introduce the main verbs and the syntax of the grammar of data **visualization** supplied by the ggplot2 package. In practice, just a single operator and a few visual marks are sufficient to effectively plot data. We also encourage the user to download the corresponding Rstudio cheat sheet [here](http://www.rstudio.com/resources/cheatsheets/) regarding data visualization with ggplot2.

* Simply stated, this grammar allows the user to **chain multiple graphical layers to construct a graph** by using the plus operator **+**, thereby improving the clarity of the code for complex graphs.


Some geometrical layers (common types of graphs):

Function | Description | Package or project
---------|-------------|-------------------
**ggplot** | map data to graph elements (axes, colors, etc...)| ggplot2
**geom_point** | plot  data as points | ggplot2
**geom_line** | join the points by lines | ggplot2
**geom_segment** | plot a segment (nematic tensor or cell bond) | ggplot2
**geom_polygon** | plot a polygon (cell contour) | ggplot2
**render_frame** | plot data onto one movie image | TissueMiner
**render_movie** | plot data onto every movie image and make a movie |TissueMiner

Some complementary scaling layers:

Function | Description | Package or project
---------|-------------|-------------------
**scale\_x\_continuous** | to control the x axis rendering | ggplot2
**scale_color_gradientn** | to use a gradient of colors when rendering the data | ggplot2


Saving a graph in the desired format (raster or vector graphics)

Function | Description | Package or project
---------|-------------|-------------------
**ggsave2** | save plots  | TissueMiner (using ggplot2)


**Example:**

**Aim:** plot the average cell area in square microns as function of time in hours from start of time-lapse recording:

* use the **ggplot()** function
* ggplot's first argument is the dataframe containing the data to be plotted
* ggplot's **aes()** function: to map the data to the system of coordinates


```{r Example ggplot, warning=FALSE, message=FALSE, fig.width=11, cache=TRUE}
# Show the first rows of the previously calculated avgCellArea data frame:
head(avgCellArea)

# Map the data to the system of coordinates using ggplot
ggplot(avgCellArea, aes(x = time_h, y = area_avg)) +
  # plot the average area as a line using geom_line
  geom_line() +
  # add a title
  ggtitle("Average cell area as function of time")
```


* save the graph with **ggsave2()**

```{r Example ggsave2, warning=FALSE, message=FALSE, fig.width=11, cache=TRUE}
# Save the plot as svg for editing in Inkscape
#ggsave2(width=14, unit="in", outputFormat="svg")
```

***

## Apply the R-grammar to visualize cells

* Cells can be rendered as polygons using the vertices ordered anti-clockewisely around each cell. Vertices ordering is performed in the automated TissueMiner workflow.
* The ordered list of vertices around each cell is stored in the *cellshapes.RData* table in the same folder as the database file.
* The **geom_polygon()** function is used to represent cells as polygons on the original image.

```{r Rendering cells and vertices load data, warning=FALSE, message=FALSE, fig.width=11,cache=TRUE}
# Load data into the 'cellshapes' variable
cellshapes <- locload(file.path(movieDir, "cellshapes.RData")) %>% print_head()

```

***

### Example 1: plot cells as polygons in the Cartesian system

```{r Rendering cells and vertices example1, warning=FALSE, message=FALSE, fig.width=11, cache=TRUE}
ggplot(cellshapes %>% filter(frame==70)) +
  # plot cells as polygons:
  geom_polygon(aes(x_pos, y_pos, group=cell_id),color="green",fill="white", size=0.3) +
  # X and Y axes must have the same scale:
  coord_equal() +
  # add a title "Pupal wing cells represented as polygons in Cartesian system" 
  ggtitle("Pupal wing cells represented as polygons in Cartesian system")
```

***

### Example 2: plot cells as polygons in the image coordinate system

```{r Rendering cells and vertices example2, warning=FALSE, message=FALSE, fig.width=11, cache=TRUE}
ggplot(cellshapes %>% filter(frame==70)) +
  geom_polygon(aes(x_pos, y_pos, group=cell_id),color="green",fill="white", size=0.3) + 
  coord_equal() +
  # In an image coordinate system, the Y-axis is pointing downwards. We flip the Y-axis:
  scale_y_continuous(trans = "reverse") +
  ggtitle("Pupal wing cells represented as polygons in image coordinate system")
```

***

### Example 3: plot cells and vertices

```{r Rendering cells and vertices example3, warning=FALSE, message=FALSE, fig.width=11, cache=TRUE}
ggplot(cellshapes %>% filter(frame==70)) +
  geom_polygon(aes(x_pos, y_pos, group=cell_id),color="green",fill="white", size=0.3) +
  # plot each vertex as a point:
  geom_point(aes(x_pos, y_pos),color="red", size=0.4) + 
  coord_equal() +
  scale_y_continuous(trans = "reverse") +
  ggtitle("Pupal wing cells and vertices")
```

***

### Example 4: overlay cells and vertices on the image

We can now overlay cells and vertices on the wing image. To do so, we built a dedicated **render_frame()** function that loads the specified frame of the time-lapse. This function takes the cell contour table and a desired frame as input variables. The **render_frame()** function returns the first layers of the graph that consists of a raster image of the wing and additional specifications such as the Y-axis flipping - **scale_y_continuous(trans = "reverse")** - and the iso-scaling of the X and Y axes - **coord_equal()**. 



```{r Rendering cells and vertices example4, warning=FALSE, message=FALSE, fig.width=11, cache=TRUE}
# Plot cells and vertices on the original image
cellshapes %>%
  # add overlay image (! connection to DB required !):
  render_frame(70) +
  geom_polygon(aes(x_pos, y_pos, group=cell_id), color="green",fill=NA, size=0.2) + 
  geom_point(aes(x_pos, y_pos),color="red", size=0.4) +
  ggtitle("Cells and vertices overlaid on the image")

```

***

## Further reading: the render_frame() function

Please, read the current definition of the **render_frame()** function at the [following location](https://github.com/mpicbg-scicomp/tissue_miner/blob/master/commons/MovieFunctions.R)

***

## Working with regions of interest (ROIs)

* By default TissueMiner creates two regions of interest:
    + **raw**: this ROI corresponds to all tracked cells contained in the DB
    + **whole_tissue**: this ROI corresponds to the largest population of cells that is visible throughout the movie. It's a subset of **raw** obtained by the lineage-browser algorithm that is part of the automated workflow of TissueMiner

* Other regions of interest can be manually defined by the user in Fiji ([see Fiji macro](https://github.com/mpicbg-scicomp/tissue_miner/blob/master/fiji_macros/draw_n_get_ROIcoord.ijm)). Of note, these additional ROIs are only taken into account if they were defined **before** running the automated workflow.  

The automated workflow includes routines to browse the cell lineage and to follow ROIs in time once defined on a given image of the time-lapse. Please note that cells in contact with the margin are discarded because the segmentation and tracking quality isn't optimum near the margin.

**Example: Visualize cells in selected ROIs**
```{r ROI definition data, warning=FALSE, message=FALSE, fig.width=11, cache=TRUE}

# Load tracked ROIs: 
lgRoiSmoothed <- locload(file.path(movieDir, "roi_bt/lgRoiSmoothed.RData")) %>%
  print_head() %>%
  filter(roi %in% c("distInterL2-L3", "distL3", "distInterL3-L4")) 

# Load cell shapes for plotting on the wing
cellshapes <- locload(file.path(movieDir, "cellshapes.RData"))

# Merge ROI with cell polygonal definition
cellshapesWithRoi <- cellshapes %>%
  dt.merge(lgRoiSmoothed, by="cell_id", allow.cartesian=T) %>%
  arrange(frame, cell_id, bond_order) ## .. because merge messes up the ordering

# Plot  ROI on the wing
render_frame(cellshapesWithRoi, 200) + 
  geom_polygon(aes(x_pos, y_pos, fill=roi, group=cell_id), alpha=0.5) +
  scale_fill_manual(values=c("distInterL2-L3"="darkgreen",
                             "distL3"="yellow",
                             "distInterL3-L4"="red"))
```
 
***

## Make videos 
**Videos are helpful to visualize the time evolution of patterns**

Here, we use a parallelized loop over all frames of the time-lapse. The well-known avconv (formerly ffmpeg) program to create videos must be installed on your computer, please, visit [http://ffmpegmac.net/](http://ffmpegmac.net/) for Mac users or simply "sudo apt-get install libav-tools" on Ubuntu-Linux. 

* To simplify the procedure of creating videos, we built a dedicated function **render_movie()** that takes a list of ggplot layers as an input argument.
* Please, read the current definition of the **render_movie()** function [here](https://github.com/mpicbg-scicomp/tissue_miner/blob/master/commons/MovieFunctions.R).
```{r ROI definition movie, warning=FALSE, message=FALSE, fig.width=11, cache=TRUE, eval=FALSE}
# Make a video of the ROIs on the wing
render_movie(cellshapesWithRoi, "bt_bhfix_peeled.mp4", list(
          geom_polygon(aes(x_pos, y_pos, fill=roi, group=cell_id),  alpha=0.5),
          scale_fill_manual(values=c("distInterL2-L3"="darkgreen",
                             "distL3"="yellow",
                             "distInterL3-L4"="red"))))
```


```{r MEM CLEAN UP 3, warning=FALSE, message=FALSE, echo=F,results='hide'}
rm(avgCellArea,cellshapes,cellshapesWithRoi,lgRoiSmoothed)
```

***

## A TissueMiner library to visualize cell dynamics

TissueMiner provide a set of tools to quantify and visualize cell dynamics at different spatial scales.
These tools are all prefixed with 'mqf' as they perform multiple queries to the pre-processed data obtained with the TissueMiner automated workflow. Their common features are:

* aggregate data from one selected movie into a dataframe for immediate visualization
* include ROI definitions
* include developmental time

Mandatory argument: a path to a selected movie directory

Optional arguments: to control selected ROIs and other parameters that are specific to some subsets of mqf functions

### Fine-grained analyses

* mqf_fg_* functions generate fine-grained data (cellular scale) that
    + are mapped to all ROIs by default
    + include bond, cell or triangle contours for plotting
    + perform an automatic scaling of nematics for an optimal display on the original image

mqf_fg_* functions | Description 
--------------|-------------
***mqf_fg_nematics_cell_elong*** | get cell elongation nematics from the DB 
***mqf_fg_unit_nematics_CD*** | calculate cell division unit nematics 
***mqf_fg_unit_nematics_T1*** | calculate cell neighbor change unit nematics 
***mqf_fg_cell_area*** | get cell area from the DB 
***mqf_fg_triangle_properties*** | get calculated triangle state properties 
***mqf_fg_bond_length*** | get bond length and positions from the DB 
***mqf_fg_cell_neighbor_count*** | calculate cell neighbor number from the DB 
***mqf_fg_dev_time*** | get developmental time from the configuration file 

***

### Coarse-grained analyses: per frame and by ROIs

* The **mqf_cg_roi_* functions**:
    + perform spatial (by ROI) and temporal (kernSize option) averaging of quantities
    + perform an automatic scaling of nematics for an optimal display on the original image

mqf_cg_roi_* functions | Description 
--------------|-------------|-------------
***mqf_cg_roi_cell_count*** | count cell number 
***mqf_cg_roi_cell_area*** | coarse-grain cell area 
***mqf_cg_roi_cell_neighbor_count*** | average cell neighbor count 
***mqf_cg_roi_polygon_class*** | average and trim cell polygon class 
***mqf_cg_roi_triangle_elong*** | coarse-grain cell elongation using triangles as a proxy 
***mqf_cg_roi_rate_CD*** | average cell division rate 
***mqf_cg_roi_rate_T2*** | average extrusion rate 
***mqf_cg_roi_rate_T1*** | average neighbor change rate 
***mqf_cg_roi_rate_isotropic_contrib*** | coarse-grain **isotropic** tissue deformation and its cellular contributions 
***mqf_cg_roi_rate_shear*** | coarse-grain **anisotropic** tissue deformation and its cellular contributions 
***mqf_cg_roi_nematics_cell_elong*** |  coarse-grain cell elongation nematics by ROI 
***mqf_cg_roi_unit_nematics_CD*** | coarse-grain division unit nematics
***mqf_cd_roi_unit_nematics_T1*** | coarse-grain neighbor change unit nematics 

***

### Coarse-grained analyses: per frame and by square-grid elements

* The **mqf_cg_grid_* functions**:
    + perform spatial (by grid element) and temporal (kernSize option) averaging of quantities
    + perform an automatic scaling of nematics for an optimal display on the original image

mqf_cg_grid_* functions | Description | Source data
--------------|-------------|-------------
***mqf_cg_grid_nematics_cell_elong*** | coarse-grain cell elongation nematics | DB
***mqf_cg_grid_unit_nematics_CD*** | coarse-grain division unit nematics | DB
***mqf_cg_grid_unit_nematics_T1*** | coarse-grain neighbor change unit nematics | DB

***

# Visualize patterned cell behaviors on the tissue

## Bond length pattern

To render cell bonds one must get bonds and their respective positions. Here, is an example in which different related tables must be joined together to pool the relevant data to be plotted:

* 3 connected tables from the relational database to be considered: **bonds**, **directed_bonds** and **vertices**
* By joining those 3 tables one can generate a new table containing the bonds with their respective vertices and length
* A bond is represented by drawing a segment between its two vertices
* Bond length is mapped onto a color gradient scale for a color-coded visualization


For the sake of clarity, we built a dedicated **mqf_fg_bond_length()** function to get bond properties along with bond positions for plotting. Please, read its definition [here](https://github.com/mpicbg-scicomp/tissue_miner/blob/master/commons/BaseQueryFunctions.R).

***

### Get and manipulate data for plotting
```{r Rendering bonds get bonds and vertices, warning=FALSE, message=FALSE, cache=TRUE, cache.lazy=FALSE}
# we use the movieDir variable defined above
bond_with_vx <- mqf_fg_bond_length(movieDir, "raw") %>% print_head()
```

***

### Plot the color-coded bond-length pattern
```{r Rendering bonds plot data, warning=FALSE, message=FALSE, fig.width=11, cache=TRUE}
bond_with_vx %>%
  render_frame(70) + 
  # bonds are represented by segments using geom_segment
  geom_segment(aes(x=x_pos.1, y=y_pos.1, xend=x_pos.2, yend=y_pos.2,
                   color=bond_length), # Here bond_length values are mapped to the color
               size=0.3, lineend="round") +
  # we overwrite the default color map by a custom rainbow palette
  scale_color_gradientn(name="bond_length",
                        colours=c("black", "blue", "green", "yellow", "red"),
                        limits=c(0,quantile(bond_with_vx$bond_length, probs = 99/100)),
                        na.value = "red") +
  ggtitle("Color-coded pattern of bond length")

```

***

### Make a video of the color-coded bond-length pattern
```{r Rendering bonds video1, warning=FALSE, message=FALSE, eval=FALSE}
# Here use the render_movie function
render_movie(bond_with_vx, "BondLengthPattern.mp4", list(
  geom_segment(aes(x=x_pos.1, y=y_pos.1,xend=x_pos.2, yend=y_pos.2,color=bond_length), 
               size=0.3, lineend="round") ,
  scale_color_gradientn(name="bond_length",
                        colours=c("black", "blue", "green", "yellow", "red"),
                        limits=c(0,quantile(bond_with_vx$bond_length, probs = 99/100)),
                        na.value = "red") # outliers are red
))
```

***

## Cell area 

### Get and manipulate the data for plotting

* Using the **TissueMiner grammar**, one can get all cell properties and cell contours in one step
```{r Cell area pattern load data, warning=FALSE, message=FALSE, fig.width=11, cache=FALSE, eval=FALSE}
cellArea <- mqf_fg_cell_area(movieDir, rois = c("raw"), cellContour = T)
```

* **Alternatively**, one can do each step separately. Here, we don't add ROI definition (faster).

```{r Cell area pattern load data stepwise, warning=FALSE, message=FALSE, fig.width=11, cache=TRUE}
# Send a SQL query to get cell area in each frame
cellArea <- dbGetQuery(db,"select cell_id, frame, area from cells where cell_id!=10000") %>%
  # add cell polygons into it
  dt.merge(locload(file.path(movieDir, "cellshapes.RData")),
           by = c("frame", "cell_id")) %>%
  # order vertices around the cell contour for ploting cells as polygons
  arrange(cell_id, frame, bond_order) 
```

***

### Plot the color-coded cell area pattern
```{r Cell area pattern rendering, warning=FALSE, message=FALSE, fig.width=11, cache=TRUE}
cellArea %>%
  render_frame(50) + 
  # we now map cell area values to a color palette: fill=area
  geom_polygon(aes(x_pos, y_pos, group=cell_id, fill=area), alpha=0.7) + 
  scale_fill_gradientn(name="area (px)",
                       colours=c("black", "blue", "green", "yellow", "red"),
                       limits=c(0,quantile(cellArea$area, probs = 99.9/100)),
                       na.value = "red") +
  ggtitle("Cell area pattern")

```

***

### Make a video of the cell area pattern
```{r Cell area pattern video, warning=FALSE, message=FALSE, eval=FALSE}
render_movie(cellArea, "CellAreaPattern.mp4", list(
  geom_polygon(aes(x_pos, y_pos, group=cell_id, fill=area)),
  scale_fill_gradientn(name="area (px)",
                       colours=c("black", "blue", "green", "yellow", "red"),
                       limits=c(0,quantile(cellArea$area, probs = 99.9/100)),
                       na.value = "red")
))

```

***

### Subset the data data by ROIs
```{r Merge ROI definition with data, warning=FALSE, message=FALSE, fig.width=11, cache=TRUE}
# We now select the blade ROI that we defined using the draw_n_get_ROIcoord.ijm Fiji macro.
cellArea <- mqf_fg_cell_area(movieDir, rois = c("raw"), cellContour = T)


whole_tissue_roi <- locload(file.path(movieDir, "roi_bt/lgRoiSmoothed.RData")) %>% 
  filter(roi=="whole_tissue") %>% print_head() 
  
cellAreaInROI <- cellArea %>%
  # A inner-join operation intersects the data
  dt.merge(whole_tissue_roi, by = "cell_id") 
```

***

### Plot the color-coded cell area pattern in the 'whole_tissue' ROI
```{r Cell area pattern in blade ROI rendering, warning=FALSE, message=FALSE, fig.width=11, cache=TRUE}
cellAreaInROI %>%
  render_frame(150) + 
  # we now map cell area values to a color palette: fill=area
  geom_polygon(aes(x_pos, y_pos, group=cell_id, fill=area)) + 
  scale_fill_gradientn(name="area (px)",
                       colours=c("black", "blue", "green", "yellow", "red"),
                       limits=c(0,quantile(cellAreaInROI$area, probs = 99.9/100)),
                       na.value = "red") +
  ggtitle("Cell area pattern in blade")

```

```{r MEM CLEAN UP 4, warning=FALSE, message=FALSE, echo=F,results='hide'}
rm(whole_tissue_roi,cellArea,cellAreaInROI)
```

***

## Cell elongation (nematic norm)

### Get and manipulate the data for plotting

* Using the **TissueMiner grammar**, one can get all cell properties and cell contours in one step

```{r Cell elongation pattern load data, warning=FALSE, message=FALSE, fig.width=11, cache=FALSE, eval=FALSE}
cellShapesElong <- mqf_fg_nematics_cell_elong(movieDir, "raw", cellContour = T)
```

* **Alternatively**, one can do each step separately. This way, one can easily filter desired cell properties

```{r Cell shape anisotropy load data, warning=FALSE, message=FALSE, fig.width=11, cache=TRUE}
# Send a SQL query to get the cell elongation tensor in each frame
cellShapesElong <- dbGetQuery(db,"select cell_id, frame, elong_xx, elong_xy from cells 
                                  where cell_id!=10000") %>%
  # calculate the norm of the elongation tensor (vectorized operation = on columns)
  mutate(norm=sqrt(elong_xx^2+elong_xy^2)) %>%
  # add cell vertices for cell rendering as polygons
  dt.merge(locload(file.path(movieDir, "cellshapes.RData")),
           by = c("frame", "cell_id")) %>%
  # order vertices around the cell contour
  arrange(cell_id, frame, bond_order) %>% print_head()
```

***

### Plot the color-coded cell elongation pattern
```{r Cell shape anisotropy plot data, warning=FALSE, message=FALSE, fig.width=11, cache=TRUE}
cellShapesElong %>%
  render_frame(70) +
  # we now map cell elongation values to a color palette: fill=elongNorm
  geom_polygon(aes(x_pos, y_pos, group=cell_id, fill=norm)) + 
  scale_fill_gradientn(name="elongation",
                       colours=c("black", "blue", "green", "yellow", "red"),
                       limits=c(0,1),
                       na.value = "red") +
  ggtitle("Cell elongation pattern")
```

***

### Make a video of the cell elongation pattern
```{r Cell elongation pattern video, warning=FALSE, message=FALSE, eval=FALSE}
render_movie(cellShapesElong, "CellElongationPattern.mp4", list(
  geom_polygon(aes(x_pos, y_pos, group=cell_id, fill=elongNorm)),
  scale_fill_gradientn(name="elongation",
                       colours=c("black", "blue", "green", "yellow", "red"),
                       limits=c(0,quantile(cellShapesElong$elongNorm, probs = 99.9/100)),
                       na.value = "red")
))

```

```{r MEM CLEAN UP 5, warning=FALSE, message=FALSE, echo=F,results='hide'}
rm(cellShapesElong)
```

***

## Cell elongation nematics (fine-grained)

### Get and manipulate the data for plotting

* Using the **TissueMiner grammar**, one can get cell elongation nematics for plotting in one step

* The **mqf_fg_nematics_cell_elong()** does the following:
    + it retrieves elongation nematics from the database. 
    + it calculates nematic angle and norm.
    + it scales nematics for display on the image (automatic scaling by default)

* Please, read the **mqf_fg_nematics_cell_elong()** definition [here](https://github.com/mpicbg-scicomp/tissue_miner/blob/master/commons/BaseQueryFunctions.R)


```{r Cell elongation nematics load data, warning=FALSE, message=FALSE, fig.width=11, cache=TRUE, eval=TRUE}
cellElongNematics <- mqf_fg_nematics_cell_elong(movieDir, "raw") %>% print_head()
```

***

### Plot the elongation nematics on each cell

* We plot nematics as segments on the original image

```{r Cell elongation nematics plot data, warning=FALSE, message=FALSE, fig.width=11, cache=TRUE}
cellElongNematics %>% 
  # crop a the image by defining squareRoi
  render_frame(120, squareRoi=rbind(c(100,400),c(1000,1500))) +
  # plot nematics as segments
  geom_segment(aes(x=x1,y=y1,xend=x2,yend=y2),
               size=1, alpha=0.7, lineend="round", color="red", na.rm=T) +
  ggtitle("Cell elongation pattern")
```

```{r MEM CLEAN UP 6, warning=FALSE, message=FALSE, echo=F,results='hide'}
rm(cellElongNematics)
```

***

## Cell elongation nematics (coarse-grained)

### Get and manipulate the data for plotting

* TissueMiner provides a **coarseGrid()** function to map cell positions to each element of a square grid. The user can find its definition [here](https://github.com/mpicbg-scicomp/tissue_miner/blob/master/commons/RoiCommons.R).

* TissueMiner provides more specific routines using this **coarseGrid()** function to average nematics in space and time. Concerning cell elongtation, the **mqf_cg_grid_nematics_cell_elong()** TissueMiner function allows the user to directly get coarse-grained nematics that are automatically scaled for display on the original image. The display factor can also be manually defined.

* The **mqf_cg_grid_nematics_cell_elong()** from the TissueMiner grammar does the following:
    + it assigns grid elements to cells and get the cell elongation tensors from the database. 
    + it averages nematics in each grid element.
    + it scales nematics for display on the image (automatic scaling by default)

* Please, read the **mqf_cg_grid_nematics_cell_elong()** definition [here](https://github.com/mpicbg-scicomp/tissue_miner/blob/master/commons/BaseQueryFunctions.R)

```{r Coarse-grained cell elongation nematics create grid, warning=FALSE, message=FALSE, cache=TRUE}
cellElongNematicsCG <- mqf_cg_grid_nematics_cell_elong(movieDir, gridSize = 96) %>%
  print_head()
```

***

### Plot coarse-grained cell elongation nematics
* We plot nematics as segments on the original image

```{r Coarse-grained cell elongation nematics plot nematics, warning=FALSE, message=FALSE, fig.width=11, cache=TRUE}

render_frame(cellElongNematicsCG, 1) +
  geom_segment(aes(x=x1, y=y1,xend=x2, yend=y2),
               size=2, lineend="round", color="red", na.rm=T)

```

***

### Make a video of the coarse-grained cell elongation nematic pattern
```{r Cell elongation nematic pattern video, warning=FALSE, message=FALSE, eval=FALSE}
render_movie(cellElongNematicsCG, "CellElongationNematicPattern.mp4", list(
  geom_segment(aes(x=x1, y=y1, xend=x2, yend=y2),
               size=2, lineend="round", color="red", na.rm=T)
))

```

```{r MEM CLEAN UP 7, warning=FALSE, message=FALSE, echo=F,results='hide'}
rm(cellElongNematicsCG)
```

***

## Cell packing geometry throughout the tissue

### Get and manipulate the data for plotting

* Using the **TissueMiner grammar**, one can get the cell neighbor count for plotting in one step

* The **mqf_fg_cell_neighbor_count()** does the following:
    + it establishes all cell-cell contact from the database. 
    + it counts the number of cell-cell contact for each cell
    + it trims the neighbor count between 4 and 8 neighbors for display
    + it retrieves cell contours for display

* Please, read the **mqf_fg_cell_neighbor_count()** definition [here](https://github.com/mpicbg-scicomp/tissue_miner/blob/master/commons/BaseQueryFunctions.R)

```{r Cell polygon class load data, warning=FALSE, message=FALSE, cache=T}
cellPolygonClass <- mqf_fg_cell_neighbor_count(movieDir, "raw", cellContour = T)
```

***

### Plot the color-coded cell neighbor number
```{r Cell polygon class plot data, warning=FALSE, message=FALSE, fig.width=11, cache=TRUE}
# Define a discrete color palette of polygon class
polygonClassColors <- c("2"="black","3"="darkgrey", "4"="green",
                        "5"="yellow", "6"="grey", "7"="blue",
                        "8"="red", "9"="purple", ">9"="black")

cellPolygonClass %>%
  render_frame(70) +
  geom_polygon(aes(x_pos, y_pos, fill=as.character(polygon_class_trimmed),
                   group=cell_id),  alpha=0.7) +
  scale_fill_manual(name="polygon class", values=polygonClassColors, drop=F) +
  ggtitle("PolygonClassPattern")

```

***

### Make a video of the cell neighbor number
```{r Cell polygon class make video, warning=FALSE, message=FALSE, eval=FALSE}
# Define a discrete color palette of polygon class
polygonClassColors <- c("2"="black","3"="darkgrey", "4"="green",
                        "5"="yellow", "6"="grey", "7"="blue",
                        "8"="red", "9"="purple", ">9"="black")

render_movie(cellPolygonClass, "CellPackingPattern.mp4", list(
  geom_polygon(aes(x_pos, y_pos, fill=as.character(polygon_class_trimmed),
                   group=cell_id),  alpha=0.7),
  scale_fill_manual(name="polygon class", values=polygonClassColors, drop=F)
))

```

```{r MEM CLEAN UP 8, warning=FALSE, message=FALSE, echo=F,results='hide'}
rm(cellPolygonClass)
```

***

## Cumulative patterns of cell divisions

### Get and manipulate the data for plotting
```{r Cell division generations get data, warning=FALSE, message=FALSE, fig.width=11, cache=TRUE}
# limit generation to a more reasonable range
genColors =c("0"="white", "1" = "red", "2"="green", "3"="cyan", ">3"="magenta")

# Send a SQL query to get the cell lineage
cellsWithLin <- dbGetQuery(db, "select cell_id, lineage_group, generation 
                                from cell_histories") %>%
  # fix a generation cut off for a more reasonable range
  mutate(generation_cutoff=ifelse(generation>3, ">3", ac(generation))) %>%
  # add cell vertices for cell rendering as polygons
  dt.merge(locload(file.path(movieDir, "cellshapes.RData")), by = "cell_id") %>%
  arrange(frame, cell_id, bond_order) %>% print_head()

```

***

### Plot color-coded cell generations
```{r Cell division generations plot data, warning=FALSE, message=FALSE, fig.width=11, cache=TRUE}
cellsWithLin %>%
  render_frame(70+80) + 
  geom_polygon(aes(x_pos, y_pos, fill=as.factor(generation_cutoff), group=cell_id), alpha=0.5) +
  scale_fill_manual(name="generation", values=genColors) +
  ggtitle("CellDivisionGeneration")

```

***

### Make a video of the time evolution of the cell generation pattern
```{r Cell division generations make video, warning=FALSE, message=FALSE, eval=FALSE}
# Define a discrete color palette of polygon class
render_movie(cellsWithLin, "CellGenerationPattern.mp4", list(
  geom_polygon(aes(x_pos, y_pos, fill=as.factor(generation_cutoff),
                   group=cell_id), alpha=0.5),
  scale_fill_manual(name="generation", values=genColors)
))

```

***


```{r MEM CLEAN UP 9, warning=FALSE, message=FALSE, echo=F,results='hide'}
rm(cellsWithLin)
```

***

## Cell division orientation (unit nematics)

### Get and manipulate the data for plotting
* Using the **TissueMiner grammar**, one can get cell division unit nematics for plotting in one step

* The **mqf_fg_unit_nematics_CD()** does the following:
    + it uses the cell center of mass of the daughter cells to calculate a unit nematic describing the division orientation
    + it scales nematics for display on the image (automatic scaling by default)

* Please, read the **mqf_fg_unit_nematics_CD()** definition [here](https://github.com/mpicbg-scicomp/tissue_miner/blob/master/commons/BaseQueryFunctions.R)

```{r Cell division nematics get data, warning=FALSE, message=FALSE, fig.width=11, cache=TRUE}
cdNematics <- mqf_fg_unit_nematics_CD(movieDir, rois = "raw", cellContour = T) %>%
  print_head()
```

***

### Plot cell division unit nematics
```{r Cell division nematics plot, warning=FALSE, message=FALSE, fig.width=11, cache=TRUE}

## Plot CD nematics on image #70 
cdNematics %>%
  render_frame(70, squareRoi=rbind(c(100,400),c(600,900))) + # x range and y range 
  geom_polygon(aes(x_pos, y_pos, group=cell_id),  alpha=0.5, fill="blue", color="grey") +
  geom_segment(aes(x=x1,y=y1,xend=x2,yend=y2),
               size=1, lineend="round", color="orange", na.rm=T) +
  ggtitle("Cell division unit nemtatic")

```

```{r MEM CLEAN UP 10, warning=FALSE, message=FALSE, echo=F,results='hide'}
rm(cdNematics)
```

***

## Coarse-grained cell division nematics with time averaging

### Get and manipulate the data for plotting

* Using the **TissueMiner grammar**, one can coarse-grain cell division unit nematics for plotting in one step

* The **mqf_cg_grid_unit_nematics_CD()** does the following:
    + it assigns grid elements to cells and calculate the cell division unit nematics
    + it averages nematics in each grid element.
    + it scales nematics for display on the image (automatic scaling by default)

* Please, read the **mqf_cg_grid_unit_nematics_CD()** definition [here](https://github.com/mpicbg-scicomp/tissue_miner/blob/master/commons/BaseQueryFunctions.R)

```{r Cell division coarse-grained nematics get data, warning=FALSE, message=FALSE, cache=TRUE}
cgCDnematicsSmooth <- mqf_cg_grid_unit_nematics_CD(movieDir, gridSize = 160, kernSize = 11)
```

***

### Plot the coarse-grained cell division nematics
```{r Cell division coarse-grained nematics plot, warning=FALSE, message=FALSE, fig.width=11, cache=TRUE}

## Plot CD nematics on image #55
cgCDnematicsSmooth %>%
  render_frame(55) + 
  geom_segment(aes(x=x1, y=y1, xend=x2, yend=y2),
               size=2, alpha=0.7, lineend="round", color="orange", na.rm=T) +
  ggtitle("Coarse-grained cell division nemtatics")

```

***

### Make a video of the pattern of the coarse-grained cell division nematics
```{r Cell division coarse-grained nematics make video, warning=FALSE, message=FALSE, eval=FALSE}
# Define a discrete color palette of polygon class
render_movie(cgCDnematicsSmooth, "cgCDnematics.mp4", list(
  geom_segment(aes(x=x1, y=y1, xend=x2, yend=y2),
               size=1, alpha=0.7, lineend="round", color="orange", na.rm=T)
))

```

```{r MEM CLEAN UP 11, warning=FALSE, message=FALSE, echo=F,results='hide'}
rm(cgCDnematicsSmooth)
```

***

## Cell neighbor changes

### Get and manipulate the data for plotting

```{r A cell-based rendering load data, warning=FALSE, message=FALSE, fig.width=11, cache=TRUE}
# Load the detected changes in topology
topoChangeSummary  <- locload(file.path(movieDir, "topochanges/topoChangeSummary.RData")) %>%
  select(cell_id, frame, num_t1_gained, num_t1_lost, num_neighbors_t)

# Filter T1 transitions and bring cellshapes for plotting
csWithTopoT1 <- locload(file.path(movieDir, "cellshapes.RData")) %>%
  dt.merge(topoChangeSummary, allow.cartesian=TRUE) %>%
  filter(num_t1_gained>0 |  num_t1_lost>0) %>%
  mutate(t1_type=ifelse(num_t1_gained>0,
                        ifelse(num_t1_lost>0, "t1 gain and loss", "t1 gain"), "t1 loss")) %>%
  print_head()

## define t1 color attribute
T1cols <- create_palette(unique(csWithTopoT1$t1_type))
```

***

### Plot cells changing neighbors
```{r A cell-based rendering plot, warning=FALSE, message=FALSE, fig.width=11, cache=TRUE}
csWithTopoT1 %>% 
  render_frame(90) + 
  geom_polygon(aes(x_pos, y_pos, group=cell_id, fill=t1_type), alpha=0.5) +
  scale_fill_manual(values = T1cols, drop = FALSE) +
  ggtitle("Cell neighbor exchanges")
```

### Make a video of cells changing neighbors
```{r A cell-based rendering make video, warning=FALSE, message=FALSE, eval=FALSE}
# Define a discrete color palette of polygon class
render_movie(csWithTopoT1, "CellNeighborChangePattern.mp4", list(
  geom_polygon(aes(x_pos, y_pos, group=cell_id, fill=t1_type), alpha=0.5),
  scale_fill_manual(values = T1cols, drop = FALSE) 
))

```

```{r MEM CLEAN UP 12, warning=FALSE, message=FALSE, echo=F,results='hide'}
rm(topoChangeSummary,csWithTopoT1)
```

***

## T1 orientation nematics (fine-grained)

### Get and manipulate the data for plotting

* Using the **TissueMiner grammar**, one can get T1 unit nematics for plotting in one step

* The **mqf_fg_unit_nematics_T1()** does the following:
    + it uses the cell center of mass of cells gaining or losing contact to calculate a unit nematic 
    + it flips (90) nematics of cells gaining contact so that all nematics would add up upon a T1 transition or cancel out if the T1 event is reversed
    + it scales nematics for display on the image (automatic scaling by default)

* Please, read the **mqf_fg_unit_nematics_T1()** definition [here](https://github.com/mpicbg-scicomp/tissue_miner/blob/master/commons/BaseQueryFunctions.R)

```{r T1 nematics, warning=FALSE, message=FALSE, cache=TRUE}
#t1nematics <- get_unit_nematics_T1(movieDir) %>% print_head()
t1nematics <- mqf_fg_unit_nematics_T1(movieDir, "raw", cellContour = T) %>% print_head()
```

***

### Plot the T1 nematics
```{r T1 nematics plot, warning=FALSE, message=FALSE, fig.width=11, cache=TRUE}
T1cols = c("gain"="red", "loss"="green", "gain_and_loss"="blue")

t1nematics %>% 
  render_frame(20, squareRoi=rbind(c(100,600),c(700,1100))) + #
  geom_polygon(aes(x_pos, y_pos, group=cell_id, fill=t1_type), alpha=0.5, color="grey") +
  scale_fill_manual(values = T1cols, drop = FALSE) +
  geom_segment(aes(x=x1,y=y1,xend=x2,yend=y2),
               size=1, alpha=0.7, lineend="round", color="red", na.rm=T)  +
  ggtitle("Cell neighbor exchanges")
```

```{r MEM CLEAN UP 13, warning=FALSE, message=FALSE, echo=F,results='hide'}
rm(t1nematics)
```

***

## T1 orientation nematics (coarse-grained)

### Get and manipulate the data for plotting

* Using the **TissueMiner grammar**, one can coarse-grain T1 unit nematics for plotting in one step

* The **mqf_cg_grid_unit_nematics_T1()** does the following:
    + it assigns grid elements to cells and calculate T1 unit nematics
    + it averages nematics in each grid element.
    + it scales nematics for display on the image (automatic scaling by default)

* Please, read the **mqf_cg_grid_unit_nematics_T1()** definition [here](https://github.com/mpicbg-scicomp/tissue_miner/blob/master/commons/BaseQueryFunctions.R)

```{r T1 coarse-grained nematics data calculation, warning=FALSE, message=FALSE, cache=TRUE}
cgT1nematics <- mqf_cg_grid_unit_nematics_T1(movieDir, rois="raw", gridSize = 160, kernSize = 11) %>%
  print_head()
```

***

### Plot the coarse-grained T1 nematics
```{r T1 coarse-grained nematics plot, warning=FALSE, message=FALSE, fig.width=11, cache=TRUE}
## Plot CD nematics on image #50
cgT1nematics %>%
  render_frame(50) + 
  geom_segment(aes(x=x1,y=y1,xend=x2,yend=y2),
               size=2, alpha=0.7, lineend="round", color="red", na.rm=T) +
  ggtitle("Coarse-grained T1 nemtatics")

```

***

### Make a video of the coarse-grained T1 nematics
```{r T1 coarse-grained nematics make video, warning=FALSE, message=FALSE, eval=FALSE}
# Define a discrete color palette of polygon class
render_movie(cgT1nematics, "cgT1nematics.mp4", list(
  geom_segment(aes(x=x1,y=y1,xend=x2,yend=y2),
               size=1, alpha=0.7, lineend="round", color="red", na.rm=T)
))

```

```{r MEM CLEAN UP 14, warning=FALSE, message=FALSE, echo=F,results='hide'}
rm(cgT1nematics)
```

***

## Comparing patterns between movies

### Movie frame synchronization
```{r synchro dev time to frame, warning=FALSE, message=FALSE, fig.width=11, cache=TRUE, cache.lazy=FALSE, eval=TRUE}
movieDirs <- file.path(movieDbBaseDir, c("WT_1","WT_2","WT_3"))
syncMovies <- multi_db_query(movieDirs, mqf_fg_cell_area, rois="raw", cellContour=T) %>%
  synchronize_frames(900) %>% print_head()
```

***

### Movie rendering
```{r render synchro movie, warning=FALSE, message=FALSE, fig.width=11, cache=TRUE, eval=F}

# use zip option of render_movie() to combine movies in Fiji

movieDir <- file.path(movieDbBaseDir, c("WT_1"))
db <- openMovieDb(movieDir)

render_movie(syncMovies %>% filter(movie=="WT_1"),
             "WT_1_CellAreaPattern.mp4", list(
  geom_polygon(aes(x_pos, y_pos, group=cell_id, fill=area)),
  scale_fill_gradientn(name="area (px)",
                       colours=c("black", "blue", "green", "yellow", "red"),
                       limits=c(0,quantile(syncMovies$area, probs = 99.9/100)),
                       na.value = "red")
), createZip = T)
dbDisconnect(db)

movieDir <- file.path(movieDbBaseDir, c("WT_2"))
db <- openMovieDb(movieDir)
render_movie(syncMovies %>% filter(movie=="WT_2"), 
             "WT_2_CellAreaPattern.mp4", list(
  geom_polygon(aes(x_pos, y_pos, group=cell_id, fill=area)),
  scale_fill_gradientn(name="area (px)",
                       colours=c("black", "blue", "green", "yellow", "red"),
                       limits=c(0,quantile(syncMovies$area, probs = 99.9/100)),
                       na.value = "red")
), createZip = T)
dbDisconnect(db) 

movieDir <- file.path(movieDbBaseDir, c("WT_3"))
db <- openMovieDb(movieDir)
render_movie(syncMovies %>% filter(movie=="WT_3"),
             "WT_3_CellAreaPattern.mp4", list(
  geom_polygon(aes(x_pos, y_pos, group=cell_id, fill=area)),
  scale_fill_gradientn(name="area (px)",
                       colours=c("black", "blue", "green", "yellow", "red"),
                       limits=c(0,quantile(syncMovies$area, probs = 99.9/100)),
                       na.value = "red")  
), createZip = T)
dbDisconnect(db) 
```

```{r MEM CLEAN UP 15, warning=FALSE, message=FALSE, echo=F,results='hide'}
rm(syncMovies)
```

***

# Comparing averaged quantities between movies and ROIs

## General principles:

**CAUTION**: Tissue orientation matters for nematic components. Indeed, nematic tensors are symmetric traceless tensors that are characterized by 2 components projected onto the x and y axis of the Cartesian system. In order to compare nematics amongst different time-lapse one has to make sure that the tissues have a similar orientation with respect to the x and y axes. In the workflow, one has the possibility to rotate the images along with the data [see Fiji macro](https://github.com/mpicbg-scicomp/tissue_miner/blob/master/fiji_macros/orient_tissue.ijm) to obtain visually comparable time-lapse. 

**CAUTION**: Cumulative quantities are strongly influenced by the developmental time. Therefore, movies must be aligned in time prior to comparison between movies. We have aligned the three WT wing movies in time by aligning the peaks of their respective average cell elongation curves as a function of time. One movie is used as a reference and time shifts are applied to other movies. These time shifts must be stored in a configuration file containing the *algnModel* table as defined [here](https://github.com/mpicbg-scicomp/tissue_miner/blob/master/config/flywing_tm_config.R).

***

### Description of the multi_db_query() function:

* To compare quantities between movies and ROi's, we built a multi_db_query() function that takes three arguments: the movies to be queried, the specific type of analysis to be performed and the ROIs to be queried. The multi_db_query() function automatically takes into account the corrected developmental time stored in the flywing_tm_config.R file.

Usage: 

* multi_db_query(movieDirectories, queryFun = mqf_cell_count, ...)

Arguments:

* movieDirectories: a list of absolute paths to the movie directories
* queryFun: a string containing the query function for a specific type of analysis to be performed. The default is mqf_cell_count, which count the number of cells per frame in each ROIs of each selected movie.
* ...: optional argument corresponding to a list of selected ROIs. By default, all existing ROIs are selected.

Returned value:

* a dataframe 

***

### Select movies and ROIs to be analyzed
```{r MultiDB-query select movie and ROIs, warning=FALSE, message=FALSE, cache=TRUE, eval=TRUE}
# Define a list of movies to compare (paths to ovie directories)
movieDirs <- file.path(movieDbBaseDir, c("WT_1",
                                         "WT_2",
                                         "WT_3"))

#  Define ROIs to be analyzed
selectedRois=c("whole_tissue","interL2-L3", "distL3")

```

***


## Averaged cell area

```{r MultiDB-query get averaged cell area, warning=FALSE, message=FALSE, cache=TRUE, eval=TRUE}
avgCellArea <- multi_db_query(movieDirs, mqf_cg_roi_cell_area, selectedRois) %>% print_head()
```

```{r MultiDB-query plot averaged cell area, warning=FALSE, message=FALSE, fig.width=11, cache=TRUE, eval=TRUE}
ggplot(avgCellArea, aes(dev_time, area.avg*(0.208^2), color=movie)) +
  geom_line() +
  ylab(expression(paste("<Cell area> [",mu,m^2,"]")))  +
  scale_color_manual(values = movieColors) +
  facet_wrap(~roi, ncol=4) +
  ggtitle("averaged cell area")
```

```{r MEM CLEAN UP 17, warning=FALSE, message=FALSE, echo=F,results='hide'}
rm(avgCellArea)
```

***

## Averaged cell elongation
```{r MultiDB-query get averaged cell elongation, warning=FALSE, message=FALSE, cache=TRUE, eval=TRUE}
# Query the DBs and calculate the norm of the cell elongation nematics for each cell
avgCellElong <- multi_db_query(movieDirs, mqf_cg_roi_nematics_cell_elong, selectedRois) %>% 
  print_head() 
```
 
```{r MultiDB-query plot averaged cell elongation, warning=FALSE, message=FALSE, fig.width=11, cache=TRUE, eval=TRUE}
ggplot(avgCellElong, aes(dev_time, norm, color=movie)) +
  geom_line() +
  ylab(expression(paste("<cell elongation norm>")))  +
  scale_color_manual(values = movieColors) +
  facet_wrap(~roi, ncol=4) +
  ggtitle("averaged cell elongation norm")
```

```{r MEM CLEAN UP 18, warning=FALSE, message=FALSE, echo=F,results='hide'}
rm(avgCellElong)
```

***

## Averaged cell neighbor number
```{r MultiDB-query get averaged cell neighbor number, warning=FALSE, message=FALSE, cache=TRUE, eval=TRUE}
avgNeighborNb <- multi_db_query(movieDirs, mqf_cg_roi_cell_neighbor_count, selectedRois) %>%
  print_head() 
```

***

```{r MultiDB-query plot averaged cell neighbor number, warning=FALSE, message=FALSE, fig.width=11, cache=TRUE, eval=TRUE}
  ggplot(avgNeighborNb, aes(dev_time, avg_num_neighbors, color=movie)) +
  geom_line() +
  ylab(expression(paste("<cell neighbor number>")))  +
  scale_color_manual(values = movieColors) +
  facet_wrap(~roi, ncol=4) +
  ggtitle("averaged cell neighbor number")
```

```{r MEM CLEAN UP 19, warning=FALSE, message=FALSE, echo=F,results='hide'}
rm(avgNeighborNb)
```

***

## Average cell neighbor number by class of polygons
```{r MultiDB-query get average polygon classes, warning=FALSE, message=FALSE, cache=F, eval=F}
avgPgClass <- multi_db_query(movieDirs, mqf_cg_roi_polygon_class, selectedRois) %>% print_head() 
```

```{r MultiDB-query plot average polygon classes, warning=FALSE, message=FALSE, fig.width=11, cache=TRUE, eval=F}
ggplot(avgPgClass, aes(dev_time, pgFreq, color=polygon_class)) +
  geom_line() +
  ylab(expression(paste("<cell neighbor number>")))  +
  # scale_color_manual(values = movieColors) +
  facet_grid(movie~roi) +
  ggtitle("averaged cell neighbor number")
```

***

## Cell division rate
```{r MultiDB-query get rate CD, warning=FALSE, message=FALSE, cache=TRUE, eval=TRUE}
CDrateByTimeIntervals <- multi_db_query(movieDirs, mqf_cg_roi_rate_CD, selectedRois) %>% 
  chunk_time_into_intervals(3600) %>%
  group_by(movie, roi,interval_mid) %>%
  summarise(avgCDrate=mean(cell_loss_rate),
            semCD=se(cell_loss_rate),
            time_sec=interval_mid[1],
            dev_time=mean(dev_time))
```

```{r MultiDB-query plot rate CD graphs, warning=FALSE, message=FALSE, fig.width=11, cache=TRUE, eval=TRUE}
ggplot(CDrateByTimeIntervals, aes(dev_time, avgCDrate, color=movie)) + 
  geom_line()+
  geom_point(size=1, color="black") +
  geom_errorbar(aes(ymin=(avgCDrate-semCD), ymax=(avgCDrate+semCD)),
                size=0.3, width=0.4, color="black") +
  ylab(expression(paste("CD rate [", cell^-1, h^-1,"]"))) + 
  scale_color_manual(values = movieColors) +
  facet_wrap(~roi) +
  ggtitle("CD rate")
```

```{r MEM CLEAN UP 20, warning=FALSE, message=FALSE, echo=F,results='hide'}
rm(CDrateByTimeIntervals)
```

***

## T1 rate
```{r MultiDB-query get rate T1, warning=FALSE, message=FALSE, cache=TRUE, eval=TRUE}
T1rate <- multi_db_query(movieDirs, mqf_cg_roi_rate_T1, selectedRois) %>% 
  chunk_time_into_intervals(deltaT = 3600) %>%
  group_by(movie, roi,interval_mid) %>%
  summarise(avgT1rate=mean(cell_topo_rate),
            semT1=se(cell_topo_rate),
            time_sec=interval_mid[1],
            dev_time=mean(dev_time))
```


```{r MultiDB-query plot rate T1 graphs, warning=FALSE, message=FALSE, fig.width=11, cache=TRUE, eval=TRUE}
ggplot(T1rate, aes(dev_time, avgT1rate, color=movie)) + 
  geom_line()+
  geom_point(size=1, color="black") +
  geom_errorbar(aes(ymin=(avgT1rate-semT1), ymax=(avgT1rate+semT1)),
                size=0.3, width=0.4, color="black") +
  facet_wrap(~roi) +
  ylab(expression(paste("T1 rate [", cell^-1, h^-1,"]"))) + 
  ylim(c(0.2,2.3)) +
  scale_color_manual(values = movieColors) +
  facet_wrap(~roi) +
  ggtitle("T1 rate")
```

```{r MEM CLEAN UP 21, warning=FALSE, message=FALSE, echo=F,results='hide'}
rm(T1rate)
```

***

## Cell division orientation (rose diag)
```{r MultiDB-query CD orientation graphs, warning=FALSE, message=FALSE, fig.width=11, cache=TRUE, eval=F}
```

***

## T1 orientation (rose diag)
```{r MultiDB-query T1 orientation , warning=FALSE, message=FALSE, fig.width=11, cache=TRUE, eval=TRUE}
movieDirs <- file.path(movieDbBaseDir, c("WT_1"))
selectedRois=c("whole_tissue")

# Get half T1 nematics and align movie start
T1Nematics <- multi_db_query(movieDirs, mqf_cg_roi_unit_nematics_T1, selectedRois) %>% 
  align_movie_start(movieDirs) %>% 
  mutate(frame=frame-closestFrame) %>% 
  group_by(movie) %>%
  mutate(maxnormByMovie=max(norm,na.rm=T)) %>% 
  group_by(movie,roi) %>%
  mutate(maxnormByRoi=max(norm,na.rm=T)) %>% print_head() 
T1Nematics$roi <- factor(T1Nematics$roi, levels=c("whole_tissue","distL3"))

```

***

```{r MultiDB-query T1 orientation graphs, warning=FALSE, message=FALSE, fig.width=11, cache=TRUE, eval=TRUE}

ggplot(T1Nematics , aes()) + 
  geom_segment(aes(x=phi, y=0, xend=phi, yend=(norm), color=dev_time),size=1, alpha=0.5) +
  geom_segment(aes(x=mod2pi(phi+pi), y=0, xend=mod2pi(phi+pi), yend=(norm),
                   color=dev_time), size=1, alpha=0.5) +
  scale_color_gradientn(name="time (hAPF)",
                        colours=c("black", "blue", "green", "yellow", "red"),
                        limits=c(min(T1Nematics$dev_time),max(T1Nematics$dev_time)),
                        na.value = "red") +
  coord_polar(start=-pi/2,direction=+1)+
  scale_x_continuous(breaks=seq(0,3*pi/2,pi/2), 
                     labels=c(expression(pi),expression(paste(pi/2," Ant")),
                              expression(0),expression(-pi/2)),
                     limits=c(0,2*pi)) +
  xlab("") +
  # scale_y_continuous(breaks=seq(0,1,0.2), limits=c(0,1)) +
  ylab("T1 nematic norm") +
  facet_wrap(~roi, ncol=4) +
  theme(#legend.justification=c(1,0),
    #legend.position=c(1,0),
    #legend.text=element_text(size=8),
    # legend.position="none",
    # legend.title=element_blank(),
    # legend.key = element_blank(),
    plot.title = element_blank(),
    # strip.text=element_blank(),
    # strip.background=element_blank(),
    panel.grid.minor=element_blank(),
    plot.margin = unit(c(0,0,0,0), "cm")) +
  ggtitle("T1 nematics - avg by roi in frame")
#ggsave2(height=6,unit="in", outputFormat="svg")

```

```{r MEM CLEAN UP 22, warning=FALSE, message=FALSE, echo=F,results='hide'}
rm(T1Nematics)
```

***

## Quantifying tissue deformation and its celluar contributions

### Isotropic deformation and its decomposition into cellular contributions
```{r MultiDB-query get isodef, warning=FALSE, message=FALSE, fig.width=11, cache=TRUE, eval=TRUE}
movieDirs <- file.path(movieDbBaseDir, c("WT_1","WT_2", "WT_3"))
selectedRois=c("whole_tissue", "distL3", "distInterL3-L4")

isoContrib <- multi_db_query(movieDirs, mqf_cg_roi_rate_isotropic_contrib, selectedRois) %>%
  filter(isoContrib!="tissue_area")

deltaT=60 # sampling (60 seconds)
avgIsoDefRateInterpolated <- isoContrib %>%
  align_movie_start(movieDirs) %>%
  mutate(min_dev_time=min(dev_time),
         max_dev_time=max(dev_time)) %>% 
  group_by(movie, roi, isoContrib) %>%
  do({
    # browser()
    with(., approx(dev_time, value.ma,
                   xout = seq(min_dev_time[1], max_dev_time[1],
                              by = deltaT/3600), method = "linear")) %>% as.df()
  }) %>%
  rename(dev_time=x, value.ma=y) %>%
  filter(!is.na(value.ma)) %>%
  ungroup() %>% mutate(movieNb=length(unique(movie))) %>% 
  group_by(roi, isoContrib, dev_time) %>%
  # Make sure that further calculations will be done on values present in all compared movies
  filter(n()==movieNb) %>% ungroup()
  
avgIsoDefRateSummary <- avgIsoDefRateInterpolated %>%
  group_by(roi, isoContrib, dev_time) %>%
  summarise(value.avg=mean(value.ma), value.sd=sd(value.ma))

```

***

```{r MultiDB-query plot isodef graphs, warning=FALSE, message=FALSE, fig.width=11, cache=TRUE, eval=TRUE}

# Plot average of iso contribution rates in 3 WT and their respective standard deviation
ggplot(avgIsoDefRateSummary, aes(dev_time, value.avg, color=isoContrib)) +
  geom_line() + 
  geom_ribbon(aes(ymin=(value.avg-value.sd), ymax=(value.avg+value.sd), fill=isoContrib),
              alpha=0.2, linetype="dotted", size=0.2) +
  xlab("Time [hAPF]")+
  # scale_x_continuous(breaks=seq(16,36, 2),limits=c(15,max(ceiling(shearRate$dev_time)))) +
  scale_x_continuous(breaks=seq(16,36, 4),limits=c(16,32)) +
  ylab(expression(paste("rate [",h^-1,"]"))) +
  #geom_hline(color="black", size=0.2) +
  #scale_y_continuous(breaks=seq(-0.2,0.2, 0.1), limit=c(-0.22, 0.22)) +
  scale_color_manual(values=isotropColors) +
  scale_fill_manual(values=isotropColors) +
  facet_wrap(~roi) +
  ggtitle("averaged rates of isotropic deformation")
```

```{r MEM CLEAN UP 23, warning=FALSE, message=FALSE, echo=F,results='hide'}
rm(avgIsoDefRateSummary)
```

***

```{r MultiDB-query get cumulative sum of area change, warning=FALSE, message=FALSE, cache=TRUE, eval=TRUE}
avgIsoDefCum <- avgIsoDefRateInterpolated %>%
  group_by(movie, roi, isoContrib) %>%
  mutate(timeInt=c(0,diff(dev_time)), value.cumsum=cumsum(value.ma*timeInt)) %>% 
  group_by(roi, isoContrib, dev_time) %>%
  summarise(cumsum.avg=mean(value.cumsum), cumsum.sd=sd(value.cumsum))
  
```

***

```{r MultiDB-query plot cumulative sum of area change, warning=FALSE, message=FALSE, fig.width=11, cache=TRUE, eval=TRUE}
ggplot(avgIsoDefCum, aes(dev_time, cumsum.avg, color=isoContrib)) +
  geom_line() +
  geom_ribbon(aes(ymin=(cumsum.avg-cumsum.sd), ymax=(cumsum.avg+cumsum.sd), fill=isoContrib),
              alpha=0.2, linetype="dotted", size=0.2) +
  xlab("Time [hAPF]")+
  ylab("cumulative sum") +
  scale_x_continuous(breaks=seq(16,36, 4),limits=c(16,32)) +
  #geom_hline(color="black", size=0.2) +
  scale_y_continuous(breaks=seq(-1, 1, 0.4), limit=c(-1, 1)) +
  scale_color_manual(values=isotropColors) +
  scale_fill_manual(values=isotropColors) +
  facet_wrap(~roi) +
  ggtitle("cumulative avg isotropic deformation")
```

```{r MEM CLEAN UP 24, warning=FALSE, message=FALSE, echo=F,results='hide'}
rm(avgIsoDefCum)
```

***

### Pure shear deformation and its decomposition into cellular contributions

#### Triangulation of the cell network


```{r MultiDB-query Triangulation of the cell network get data, warning=FALSE, message=FALSE, cache=T, cache.lazy=FALSE,eval=TRUE}
triProperties <- mqf_fg_triangle_properties(movieDir, "raw", triContour = T) %>%
  print_head()

```

***

```{r MultiDB-query Triangulation of the cell network plot, warning=FALSE, message=FALSE, fig.width=11, cache=T, eval=TRUE}
triProperties %>%
  render_frame(40) +
  geom_polygon(aes(x_pos, y_pos, group=tri_id),
               fill="yellow", color="black", alpha=0.9, size=0.1)
```

```{r MEM CLEAN UP 25, warning=FALSE, message=FALSE, echo=F,results='hide'}
rm(triProperties)
```

***

#### Pure shear deformation
```{r MultiDB-query get shear deformation, warning=FALSE, message=FALSE, fig.width=11, cache=TRUE, eval=TRUE}
shearData <- multi_db_query(movieDirs, mqf_cg_roi_rate_shear, selectedRois)

shearRateSlim <- subset(shearData, (tensor=="CEwithCT" | tensor=="correlationEffects" |
                                      tensor=="nu" | tensor=="ShearT1" | 
                                      tensor=="ShearT2" | tensor=="ShearCD"))
shearRateSlim$tensor <- factor(shearRateSlim$tensor, 
                               levels=c("ShearCD", "CEwithCT", "correlationEffects",
                                        "nu", "ShearT1", "ShearT2"),
                               labels=c("cell_division", "cell_elongation_change",
                                        "correlation_effects","total_shear","T1", "T2"))

deltaT=60
shearRateInterpolated <- shearRateSlim %>% #filter(movie=="WT_25deg_111102") %>%
  align_movie_start(movieDirs) %>%
  select(-c(xy,yx,yy,xy.ma,yx.ma,yy.ma, TimeInt.ma,
            phi, norm,time_sec,timeInt_sec,closestFrame,time_shift)) %>%
  mutate(min_dev_time=min(dev_time),
         max_dev_time=max(dev_time)) %>% 
  group_by(movie, roi, tensor) %>%
  do({
    with(., approx(dev_time, xx.ma,
                   xout = seq(min_dev_time[1], max_dev_time[1],
                              by = deltaT/3600), method = "linear")) %>% as.df()
  }) %>%
  rename(dev_time=x, XX=y) %>% 
  filter(!is.na(XX)) %>%
  ungroup() %>% mutate(movieNb=length(unique(movie))) %>% 
  group_by(roi, tensor, dev_time) %>%
  # Make sure that further calculations will be done on values present in all compared movies
  filter(n()==movieNb) %>% ungroup()


shearRateSummary <- shearRateInterpolated %>%
  group_by(roi, tensor, dev_time) %>%
  summarise(xx.avg=mean(XX), xx.sd=sd(XX))

```

***

```{r MultiDB-query plot shear deformation, warning=FALSE, message=FALSE, fig.width=11, cache=TRUE, eval=TRUE}

# Plot avg and standard deviation for each tensor among 3 WT
ggplot(shearRateSummary, aes(dev_time,xx.avg*100, color=tensor)) +
  geom_line() + 
  geom_ribbon(aes(ymin=100*(xx.avg-xx.sd), ymax=100*(xx.avg+xx.sd), fill=tensor),
              alpha=0.2, linetype="dotted", size=0.2) +
  xlab("Time [hAPF]")+
  # scale_x_continuous(breaks=seq(16,36, 2),limits=c(15,max(ceiling(shearRate$dev_time)))) +
  scale_x_continuous(breaks=seq(16,36, 2),limits=c(16,34)) +
  ylab(expression(paste("shear rate xx [",10^-2,h^-1,"]"))) +
  # scale_y_continuous(breaks=seq(-6,8, 2), limit=c(-4.5, 7.5)) +
  scale_color_manual(values=shearColors) +
  scale_fill_manual(values=shearColors) +
  facet_wrap(~roi) +
  ggtitle("shear decomposition")
```

***

```{r MultiDB-query get cumulative sum of anisotropic deformation, warning=FALSE, message=FALSE, cache=TRUE, eval=TRUE}

shearCumSumSummary <- shearRateInterpolated %>%
  group_by(movie, roi, tensor) %>%
  mutate(timeInt=c(0,diff(dev_time)),cumSum_xx=cumsum(XX*timeInt)) %>%
  group_by(roi, tensor, dev_time) %>%
  summarise(xxCumSum.avg=mean(cumSum_xx, na.rm = F), xxCumSum.sd=sd(cumSum_xx, na.rm = F))

```

***

```{r MultiDB-query plot cumulative sum of anisotropic deformation, warning=FALSE, message=FALSE, fig.width=11, cache=TRUE, eval=TRUE}

ggplot(shearCumSumSummary, aes(dev_time,xxCumSum.avg, color=tensor)) +
  geom_line() + 
  geom_ribbon(aes(ymin=(xxCumSum.avg-xxCumSum.sd), ymax=(xxCumSum.avg+xxCumSum.sd), fill=tensor),
              alpha=0.2, linetype="dotted", size=0.2) +
  xlab("Time [hAPF]")+
  # scale_x_continuous(breaks=seq(16,36, 2),limits=c(15,max(ceiling(shearRate$dev_time)))) +
  scale_x_continuous(breaks=seq(16,36, 2),limits=c(16,34)) +#31.3
  # scale_y_continuous(breaks=seq(-0.2,0.3, 0.1), limit=c(-0.2, 0.3)) +
  ylab(expression(paste("cumulative PD shear"))) +
  #geom_hline(color="black", size=0.2) +
  scale_color_manual(values=shearColors) +
  scale_fill_manual(values=shearColors) +
  facet_wrap(~roi) +
  ggtitle("cumulative shear decomposition")

```







