---
output: 
  html_document: 
    keep_md: yes
---

# Cell elongation analysis

Cell elongation is stored in the database. It is therefore sufficient to build the database only.


### 1. Make a movie of cell elongation magnitude pattern plotted on the tissue

* Copy-paste the following commands in the terminal:

```
sm make_db 
cell_elongation_magnitude_pattern.R . output_analysis
```

```{r cell_elongation_magnitude_pattern, warning=FALSE, message=FALSE, fig.width=11, cache=FALSE, eval=TRUE, echo=FALSE, results='hide'}
movieDir="/Users/retourna/MovieDB/demo"

scriptsDir=Sys.getenv("TM_HOME")
if(is.na(file.info(scriptsDir)$isdir)){
  stop(paste("TM_HOME not correctly defined (",scriptsDir ,")"))
}

source(file.path(scriptsDir, "commons/TMCommons.R"))
source(file.path(scriptsDir, "commons/BaseQueryFunctions.R"))
source(file.path(scriptsDir, "config/default_config.R"))

cellElongNematics <- mqf_fg_nematics_cell_elong(movieDir, rois = c("raw"), cellContour = T)

db <- openMovieDb(movieDir)

cellElongNematics %>%
  render_frame(10) +
  # we now map cell elongation values to a color palette: fill=elongNorm
  geom_polygon(aes(x_pos, y_pos, group=cell_id, fill=norm)) + 
  scale_fill_gradientn(name="elongation",
                       colours=c("black", "blue", "green", "yellow", "red"),
                       limits=c(0,1),
                       na.value = "red") +
  ggtitle("cell_elongation_pattern")


```

[How to look at the results ?](../tm_qs_example_data.md#4-look-at-the-results) **|** 
[Back to tutorial list](../tm_qs_example_data.md#3-select-the-analysis-you-are-interested-in)


### 2. Make a movie of coarse-grained cell elongation nematics plotted on the tissue
* Copy-paste the following commands in the terminal:

```
sm make_db 
cell_elongation_nematic_cg_pattern.R . output_analysis
```

```{r cg_cell_elongation_nematic_pattern, warning=FALSE, message=FALSE, fig.width=11, cache=FALSE, eval=TRUE, echo=FALSE, results='hide'}
cellElongNematicsCG <- mqf_cg_grid_nematics_cell_elong(movieDir, gridSize = 96) 
render_frame(cellElongNematicsCG, 1) +
  geom_segment(aes(x=x1, y=y1,xend=x2, yend=y2),
               size=2, lineend="round", color="red", na.rm=T)

```

[How to look at the results ?](../tm_qs_example_data.md#4-look-at-the-results) **|** 
[Back to tutorial list](../tm_qs_example_data.md#3-select-the-analysis-you-are-interested-in)

### 3. Plot cell elongation distrubution and averages
* Copy-paste the following commands in the terminal:

```
sm make_db 
cell_elongation_graphs.R . output_analysis
```

```{r cell_elongation_graphs, warning=FALSE, message=FALSE, fig.width=7, cache=FALSE, eval=TRUE, echo=FALSE,results='hide'}

mqf_cg_roi_nematics_cell_elong(movieDir, rois = c("raw")) %>% 
  ggplot(aes(dev_time, norm, color=movie)) +
  geom_line() +
  xlab("Time [h]") +
  ylab("<Cell elongation norm>")  +
  facet_wrap(~roi) +
  ggtitle("averaged_cell_elongation")

mqf_fg_nematics_cell_elong(movieDir, rois = c("raw")) %>% 
  ggplot(aes(norm, fill=movie)) +
  geom_histogram(color="white") +
  xlab("Cell elongation norm")  +
  facet_wrap(~roi) +
  ggtitle("cell_elongation_distribution")

mqf_fg_nematics_cell_elong(movieDir, rois = c("raw")) %>%
  ggplot(aes(movie,norm, fill=movie)) +
  geom_boxplot() +
  ylab("Cell elongation norm")  +
  facet_wrap(~roi) +
  ggtitle("cell_elongation_boxplot")

mqf_fg_nematics_cell_elong(movieDir, rois = c("raw")) %>%
  ggplot(aes(movie,norm, fill=movie)) +
  geom_violin() +
  ylab("Cell elongation norm")  +
  facet_wrap(~roi) +
  ggtitle("cell_elongation_violin")
```

[How to look at the results ?](../tm_qs_example_data.md#4-look-at-the-results) **|** 
[Back to tutorial list](../tm_qs_example_data.md#3-select-the-analysis-you-are-interested-in)

### 4. For further details

* compare multiple movies and ROI's, see [TM R User Manual](https://mpicbg-scicomp.github.io/tissue_miner/tm_tutorial/R-tutorial.html#comparing-averaged-quantities-between-movies-and-rois)
